---
title: "tidycensus Tutorial"
author: "Alex Brasch"
date: "9/23/2020"
output:
  html_document:
    includes:
      before_body: header.html
      after_body: footer.html
    code_folding: hide
    highlight: zenburn
    self_contained: yes
    theme: darkly
    toc: yes
    toc_depth: '2'
    toc_float:
      collapsed: no
      toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
editor_options:
  chunk_output_type: console
always_allow_html: yes
---

```{css, echo = FALSE}
  pre:not([class]) {
    color: #333333;
    background-color: #cccccc;
  }

blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 5px solid #eee;
}
```

```{r setup, echo=FALSE, warning=FALSE, error=FALSE, results='hide', message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, error=FALSE, results='hide', message=FALSE}
# Require the pacman package to easily load all necessary packages
if(!require(pacman)){install.packages("pacman");library(pacman)}
suppressPackageStartupMessages(p_load(
  tidyverse,
  tidycensus,
  tigris,
  sf,
  rgdal,
  readxl,
  writexl,
  xlsx,
  arcgisbinding,
  mapview,
  leaflet,
  kableExtra,
  janitor,
  extrafont))

# census_api_key("63beeb125c35332271ff1882d7935ba46c15a7d4")

# Set options
options(tigris_class = "sf", tigris_use_cache = T) # Return an object of class sf via tigris; do not cache Census shapefile downloads
options(stringsAsFactors = F) # R often uses a concept of factors to re-encode strings. This can be too early and too aggressive. Sometimes a string is just a string. To avoid problems delay re-encoding of strings by using stringsAsFactors = FALSE when creating data.frames.
options(dplyr.width = Inf) # In response to “Variables not shown” in dplyr; overrides the width of columns that gets printed out (i.e., to display all columns from df)
options(survey.replicates.mse = T) # options("survey.replicates.mse") controls the default in svrepdesign and as.svrepdesign for computing variances. When options("survey.replicates.mse") is TRUE, the default is to create replicate weight designs that compute variances centered at the point estimate, rather than at the mean of the replicates.
options(scipen = 999) # 'scipen': integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation. Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than 'scipen' digits wider.
options(datatable.fread.datatable=F) # Similar to read.table but faster and more convenient. All controls such as sep, colClasses and nrows are automatically detected. bit64::integer64 types are also detected and read directly without needing to read as character before converting.
```

# Reference & Startup

The documentation and examples within this tutorial were gleaned from the following resources:

[tidycensus](https://walkerke.github.io/tidycensus/ "tidycensus")  
[tigris](https://www.rdocumentation.org/packages/tigris/versions/1.0 "tigris")  
[tidyverse](https://www.tidyverse.org "tidyverse")  
[Census Developers](https://www.census.gov/developers/ "Census Developers")  
[Census Geography Program](https://www.census.gov/programs-surveys/geography.html "Census Geography Program")  
[Leaflet for R](https://rstudio.github.io/leaflet/ "Leaflet for R")

`tidycensus` is an R package that allows users to interface with the U.S. Census Bureau’s decennial census and American Community Survey (ACS) APIs, in  order to retrieve demographic and economic data for specified geographies. As noted by its author, Kyle Walker, `tidycensus` "returns tidyverse-ready data frames of selected variables, and the option to include simple feature (sf) geometries".

The main function of the decennial census is to provide counts of people for the purpose of congressional apportionment and federal funding allocation, while the primary purpose of the ACS is to measure the changing social and economic characteristics of the U.S. population, including education, housing, jobs, and more. Due to their differing purposes, variables that exist in the decennial census may not be included in the ACS and vice versa. The same is true between ACS years, due to the evolving nature of the surveys (i.e., questions may be added, removed, or revised).

The decennial census is an enumeration, meaning it aims to count the entire population of the country (at the location where each person usually lives). The decennial census asks a relatively small set of questions of people in homes and group living situations, including how many people live or stay in each home, as well as the sex, age, and race of each person.

ACS data differ from decennial census data in that ACS data are based on an annual sample of households, rather than a complete enumeration. In turn, ACS data points are estimates characterized by a margin of error (MOE). `tidycensus` will always return the estimate and MOE for any requested variables. When requesting ACS data with `tidycensus`, it is not necessary to specify the "E" or "M" suffix for a variable name. Available survey types include ACS 5-year estimates (`acs5`) or ACS 1-year estimates (`acs1`). Note that the latter is only available for geographies with populations of 65,000 and greater.

## Install and load packages

To get started, load the `tidycensus` and `tidyverse` packages. Additional packages used within this RMarkdown file include  tigris, readxl, writexl, arcgisbinding, leaflet, kableExtra, janitor, and extrafont.  
A Census API key is also required, which can be obtained from http://api.census.gov/data/key_signup.html. Entry of the key using the `census_api_key` function only needs to occur once (i.e., it is tied to RStudio, rather than a single R script or Markdown file).

```{r class.source = 'fold-show'}
# Uncomment as-needed
# install.packages("tidyverse")
# install.packages("tidycensus")

# library(tidyverse)
# library(tidycensus)

# census_api_key("YOUR API KEY GOES HERE")
```

# Usage

The following section contains examples of using the `tidycensus` package and Census API to retrieve, prepare, reshape, and blend demographic data for analysis and visualization. To save on processing time and avoid local memory limitations, data-intensive code chunks have been commented out (e.g., retrieving large amounts of census blocks using the `tidycensus` or `tigris` packages). Readers can view the underlying code while the input data is read-in as part of the R Project's local data.

## Review Variables

```{r class.source = 'fold-show'}
# Load the 2010 decennial census variables
DC2010_sf1 <- load_variables(2010, "sf1", cache = TRUE)
# View(DC2010_sf1)

# Load the 2014-2018 5-year ACS variables
ACS2018_acs5 <- load_variables(2018, "acs5", cache = TRUE)
# View(ACS2018_acs5)
```


```{r}
# View a portion of the 2010 decennial census table
DC2010_sf1 %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# View a portion of the 2018 ACS table
ACS2018_acs5 %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Example 1 - Decennial Census

Retrieve the 2010 decennial census total population for all U.S. states.

- Define a geography (e.g., states, counties, tracts)
- Define a single variable

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve the 2010 decennial census total population for all U.S. states.
st_2010_dc <- tidycensus::get_decennial( # API call
  geography = 'state', # Specify geography
  variables = 'P001001', # Specify variable
  year = 2010, # Specify year
  geometry = FALSE, # Include or omit spatial geometry
  output = 'wide', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

```{r}
# View data
st_2010_dc %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Example 2 - Decennial Census

Retrieve 2010 decennial census variables and geometries for a select set of a specified geography.

- Define a geography and specify a subset
- Define and name multiple variables
- Retrieve geometries

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve 2010 decennial census variables and geometries for a select set of a specified geography.
co_2010_dc <- tidycensus::get_decennial( # API call
  geography = 'county', # Specify geography
  state = 'OR', # Specify state abbreviation
  county = c('Multnomah', 'Washington', 'Clackamas'), # Specify a list of counties
  variables = c(pop_tot = 'P001001', # Specify and name multiple variables
                pop_sex_m_tot = 'P012002',
                pop_sex_f_tot = 'P012026'),
  year = 2010, # Specify year
  geometry = TRUE, # Include or omit spatial geometry
  output = 'wide', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

```{r}
# View data
co_2010_dc %>% st_set_geometry(NULL) %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Example 3 - ACS

Retrieve 2018 ACS 5-year variables and geometries for a select set of a specified geography.

- Define a geography and specify a subset
- Define and name multiple variables
- Retrieve geometries, specifically the TIGER/Line shapefiles

Concerning geometries, `tidycensus` used the geographic coordinate system NAD 1983 (EPSG: 4269), which is the default for Census spatial data files. `tidycensus` uses the Census cartographic boundary shapefiles for faster processing; if you prefer the TIGER/Line shapefiles (i.e., Topologically Integrated Geographic Encoding and Referencing), set `cb = FALSE` in the function call. Per Census documentation, the cartographic boundary files are simplified representations of selected geographic areas from the Census Bureau’s Master Address File (MAF)/TIGER geographic database. These boundary files are specifically designed for small scale thematic mapping. When possible, generalization is performed with intent to maintain the hierarchical relationships among geographies and to maintain the alignment of geographies within a file set for a given year. To improve the appearance of shapes, areas are represented with fewer vertices than detailed TIGER/Line equivalents. Some small holes or discontiguous parts of areas are not included in generalized files. Generalized boundary files are clipped to a simplified version of the U.S. outline. As a result, some off-shore areas may be excluded from the generalized files. Consult this [TIGER Data Products Guide](https://www.census.gov/programs-surveys/geography/guidance/tiger-data-products-guide.html "TIGER Data Products Guide") to determine which file type is best for your purposes.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve 2018 ACS 5-year variables and geometries for a select set of a specified geography.
co_2018_acs5 <- tidycensus::get_acs( # API call
  geography = 'county', # Specify geography
  state = 'OR', # Specify state abbreviation
  county = c('Multnomah', 'Washington', 'Clackamas'), # Specify a list of counties
  variables = c(hh_medinc_total = 'B19013_001', # Specify and name multiple variables
                hh_foodst_total = 'B22003_001', 
                hh_foodst_rec = 'B22003_002'),
  year = 2018, # Specify year
  survey = 'acs5', # Specify survey type
  geometry = TRUE, # Include or omit spatial geometry
  cb = FALSE, # Specify cartographic boundary files or TIGER/Line shapefiles
  output = 'wide', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

```{r}
# View data
co_2018_acs5 %>% st_set_geometry(NULL) %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

Concerning the structure of the data frame, a wide format contains a single row for each observation with many columns representing all variables (human-readable), while a long/tidy format contains many rows per observation (assuming more than one variable) with name-value pairs for each variable and associated value (machine-readable. For more details, see Hadley Wickham's seminal paper [Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf "Tidy Data").

## Example 4 - ACS

Retrieve 2015 ACS 1-year variables and geometries for a select set of a specified geography.

- Define a geography and specify a subset
- Define and name multiple variables
- Retrieve geometries, specifically the cartographic boundaries
- Output in long/tidy format

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve 2015 ACS 1-year variables and geometries for a select set of a specified geography.
co_2015_acs1 <- tidycensus::get_acs( # API call
  geography = 'county', # Specify geography
  state = 'OR', # Specify state abbreviation
  county = c('Multnomah', 'Washington', 'Clackamas'), # Specify a list of counties
  variables = c(hh_medinc_total = 'B19013_001', # Specify and name multiple variables
                hh_foodst_total = 'B22003_001', 
                hh_foodst_rec = 'B22003_002'),
  year = 2015, # Specify year
  survey = 'acs1', # Specify survey type
  geometry = T, # Include or omit spatial geometry
  # By default, cartographic boundary files are used
  output = 'tidy', # Set to wide or tidy/long format
  cache_table = T # Cache the table so it can be called quicker in the future, or not to save memory
)
```

Compare the structure of the data sets.

```{r}
# View the 2018 ACS 5-year table in wide format
co_2018_acs5 %>% st_set_geometry(NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# View the 2015 ACS 1-year table in long/tidy format
co_2015_acs1 %>% st_set_geometry(NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Example 5 - ACS

Retrieve 2018 ACS 5-year variables and geometries for a geography within a larger geography (e.g., all counties within a state).

- Create a vector of all members of a specified geography (e.g., all counties in Oregon)
- Define and name multiple variables
- Retrieve geometries, specifically the cartographic boundaries

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve 2018 ACS 5-year variables and geometries for all members of a specified geography within a larger geography (e.g., all counties within a state).
county_vector <- tidycensus::fips_codes %>% # Retrieve table of all state/county names and associated FIPS codes
  filter(state_name == "Oregon") %>% # Specify state
  select(county_code, county) # Maintain only the county code and name variables

# Retrieve variables and geometries for all counties within the vector
co_OR_2018_acs5 <- tidycensus::get_acs( # API call
  geography = 'county', # Specify geography
  state = 'OR', # Specify state abbreviation
  county = county_vector$county_code, # Vector of all counties
  variables = c(hh_medinc_total = 'B19013_001', # Specify and name multiple variables
                hh_foodst_total = 'B22003_001', 
                hh_foodst_rec = 'B22003_002'),
  year = 2018, # Specify year
  survey = 'acs5', # Specify survey type
  geometry = TRUE, # Include or omit spatial geometry
  cb = TRUE, # Specify cartographic boundary files or TIGER/Line shapefiles
  output = 'tidy', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

```{r}
# View data
co_OR_2018_acs5 %>% st_set_geometry(NULL) %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Example 6 - ACS

Retrieve a complete table of 2018 ACS 5-year variables and geometries for a select set of a specified geography.

- Define a geography and specify a subset
- Define a table of variables
- Retrieve geometries, specifically the TIGER/Line shapefiles

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve a complete table of 2018 ACS 5-year variables and geometries for a select set of a specified geography.
co_2018_acs5_income <- tidycensus::get_acs( # API call
  geography = 'county', # Specify geography
  state = 'OR', # Specify state abbreviation
  county = c('Multnomah', 'Washington', 'Clackamas'), # Specify a list of counties
  table = "B19001",
  year = 2018, # Specify year
  survey = 'acs5', # Specify survey type
  geometry = TRUE, # Include or omit spatial geometry
  cb = FALSE, # Specify cartographic boundary files or TIGER/Line shapefiles
  output = 'wide', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

## Example 7 - ACS

At a time in 2019, retrieving all decennial census block group data for a specified state or county generates an [error](https://github.com/walkerke/tidycensus/issues/193 "error"). This has since been resolved, but it provides a good example of how smaller nested geographies can be aggregated to larger geographies (e.g., blocks to block groups).

**This works now...**

Retrieve block group data for an entire county.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
bg_2010_dc <- tidycensus::get_decennial( # API call
  geography = 'block group', # Specify geography for which to retrieve data
  state = 'WA', # Specify state abbreviation
  county = 'King', # Specify county
  variables = 'P001001', # Specify variable
  year = 2010, # Specify year
  geometry = FALSE, # Include or omit spatial geometry
  output = 'wide', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

**But if it didn't...**

Retrieve block data for an entire county.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve decennial census variables for all blocks
bl_2010_dc <- tidycensus::get_decennial(
  geography = 'block', # Specify statistical area for which to retrieve data
  state = 'WA', # Specify state abbreviation
  county = 'King', # Specify county
  variables = 'P001001',
  year = 2010, # Specify year
  geometry = FALSE, # Include or omit spatial geometry
  output = 'wide', # Set to wide or tidy/long format
  cache_table = TRUE # Cache the table so it can be called quicker in the future, or not to save memory
)
```

Aggregate data to block groups via creation of the block group GEOID by removing the last 3 characters in the block GEOID and grouping by/summarizing to block groups.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
bl_2010_dc_to_bg <- bl_2010_dc %>% 
  mutate(GEOID_BG = stringr::str_sub(GEOID, 1, -4)) %>% # Create BG GEOID
  select(GEOID_BG, 1:length(.), -GEOID, -NAME) %>% # Reorder/remove columns
  rename(GEOID = GEOID_BG) %>% # Rename GEOID
  gather(variable, est, 2:length(.)) %>% # Reshape from wide to long form by creating name-value pairs for all variables
  group_by(GEOID, variable) %>% # Group by GEOID and variable names
  summarize(est = sum(est, na.rm = T)) %>% # Sum values per group
  ungroup() %>% # Ungroup
  spread(variable, est) # Reshape from long to wide form by extending values into their own columns
```

Compare the data sets.

```{r}
bgbl_2010_dc <- bg_2010_dc %>%
  inner_join(bl_2010_dc_to_bg, by = 'GEOID')
```

```{r}
# View the comparison data set
bgbl_2010_dc %>% head() %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Example 8 - ACS and tigris

In some cases, you may want retrieve the tabular and spatial data separately (to avoid very large data sets during analysis) and join the data sets after analysis. In those cases, `tidycensus` can be used in combination with `tigris`, which is an R package that allows users to directly download and use TIGER/Line shapefiles from the US Census Bureau.

Retrieve geometries for a specified geography for a geography within a larger geography (e.g., all tracts within a county).

- Create a vector of variables
- Create a vector of geographies
- Use `tidyensus` to retrieve tabular data
- Use `tigris` to retrieve spatial geometries
- Join the tabular and spatial data

Retrieve variables for a specified geography via the `tidycensus` package.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Create a vector of variables for use in the `tidycensus` call.
DC2010_sf1_var <- c(pop_sex_total = 'P012001',
                pop_sex_m_tot = 'P012002',
                pop_sex_m_00_04 = 'P012003',
                pop_sex_m_05_09 = 'P012004',
                pop_sex_m_10_14 = 'P012005',
                pop_sex_m_15_17 = 'P012006',
                pop_sex_m_18_19 = 'P012007',
                pop_sex_f_tot = 'P012026',
                pop_sex_f_00_04 = 'P012027',
                pop_sex_f_05_09 = 'P012028', 
                pop_sex_f_10_14 = 'P012029', 
                pop_sex_f_15_17 = 'P012030', 
                pop_sex_f_18_19 = 'P012031')

# Retrieve variables for specified geography
tr_2010_dc <- tidycensus::get_decennial( # API call
geography = 'tract', # Specify geography
variables = DC2010_sf1_var, # Vector of variables
state = 'OR', # Specify state abbreviation
county = county_vector$county_code, # Vector of all counties
year = 2010, # Specify year
geometry = F, # Include or omit spatial geometry
output = 'wide', # Set to wide or long/tidy format
cache_table = T # Cache the table so it can be called quicker in the future
)
```

Retrieve geometries for a specified geography via the `tigris` package. 

By default `tigris` retrieves the most recent vintage of a data set, so specify a different year if-needed. The default coordinate reference system (CRS) is NAD83 (EPSG 4269 https://spatialreference.org/ref/epsg/nad83/). The default type of geometry is TIGER/Line file. If cb is set to TRUE, `tigris` will download a generalized (1:500k) set of geometries.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# A call of gc causes a garbage collection to take place. The primary purpose of calling gc is for the report on memory usage. Use this before and after a data-heavy processing task.
gc() 
tr_2010_dc_shp <- tigris::tracts("OR", year = 2010, cb = FALSE) # Specify state abbreviation, year, and geometry type
gc()
```

Join the variables to the geometries.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
tr_2010_dc_geo <- tr_2010_dc_shp %>%
  inner_join(tr_2010_dc, by = c("GEOID10" = "GEOID")) # Join based on GEOID

class(tr_2010_dc_geo) # Note the object's class
st_crs(tr_2010_dc_geo) # Note the object's coordinate system
```

Note that the resulting object's class is dependent on the join order. The left side's class takes priority; therefore, in the above, the attributes (right side) are being joined to the geometries (left side), so the resulting object class is `sf` (simple feature). If the order is flipped (below) and the geometries (right side) are joined to the attributes (let side), the object class is not `sf`. To make it so, add `%>% st_as_sf()`

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
tr_2010_dc_geo_REVERSE1 <- tr_2010_dc %>%
  inner_join(tr_2010_dc_shp, by = c("GEOID" = "GEOID10"))

class(tr_2010_dc_geo_REVERSE1)
st_crs(tr_2010_dc_geo_REVERSE1)

tr_2010_dc_geo_REVERSE2 <- tr_2010_dc %>%
  inner_join(tr_2010_dc_shp, by = c("GEOID" = "GEOID10")) %>%
  st_as_sf()

class(tr_2010_dc_geo_REVERSE2)
st_crs(tr_2010_dc_geo_REVERSE2)
```

# Visualization

`ggplot2` is a data visualization package that is part of the `tidyverse`.

Create a plot of 2010 decennial census state populations.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
st_2010_dc %>%
  ggplot(aes(x = P001001, y = reorder(NAME, P001001,))) + # Set the x-axis to the variable and the y-axis to the observation (sort in descending order)
  ggtitle("2010 State Population") + # Add a title
  xlab("Total Population") + # Name the x-axis
  ylab("State Name") + # Name the y-axis
  geom_point() # Add point geometry
```

Create a plot of 2014-2018 ACS 5-year estimates and MOEs for all Oregon counties.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
co_OR_2018_acs5 %>%
  mutate(NAME = gsub(" County, Oregon", "", NAME)) %>% # Rename unnecessary portion of county name
  filter(variable == 'hh_medinc_total') %>% # Query a single variable
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +  # Set the x-axis to the variable and the y-axis to the observation (and sort)
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + # Create error bars
  geom_point(color = "red", size = 3) + # Symbolize the points
  labs(title = "Median Household Income by County", # Add title
       subtitle = "2014-2018 (5-year) American Community Survey", # Add subtitle
       y = "County Name", # Label y-axis
       x = "ACS estimate (bars represent margin of error)") # Label x-axis
```

Create a choropleth map of a single variable for a geography within a single county.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Retrieve variables and geographies for census block groups in Multnomah County, Oregon
bg_Mult_2018_acs5 <- tidycensus::get_acs( # API call
  geography = 'block group', # Specify geography
  state = 'OR', # Specify state name
  county = c('Multnomah'), # Specify county name
  variables = c(hh_medinc_total = 'B19013_001'),
  year = 2018, # Specify year
  survey = 'acs5', # Specify survey type
  geometry = T, # Include or omit spatial geometry
  output = 'tidy', # Set to wide or tidy/long format
  cache_table = T # Cache the table so it can be called quicker in the future, or not to save memory
)

# Create choropleth map
bg_Mult_2018_acs5 %>%
  ggplot(aes(fill = estimate)) + # Specify the aesthetic as fill based on the estimate values
  geom_sf(color = NA) + # Specify outline color
  coord_sf(crs = 26910) + # Specify coordinate system/projection (e.g., UTM10N)
  ggtitle("Multnomah County Median Household Income", subtitle = "2014-2018 (5-year) ACS Data") + # Specify title and subtitle
  scale_fill_viridis_c(option = "magma") # Specify fill symbology
```

Created faceted choropleths maps of multiple variables across geographies within a single county.

As mentioned by Kyle Walker in his `tidycensus` tutorial, "one of the most powerful features of `ggplot2` is its support for small multiples, which works very well with the tidy data format returned by `tidycensus`. Many Census and ACS variables return _counts_, which are generally inappropriate for choropleth mapping. In turn, `get_decennial` and `get_acs` have an optional argument, `summary_var`, that can work as a multi-group denominator when appropriate." For example, view the racial/ethnic population distribution within a given county.

```{r class.source = 'fold-show', results='hide', message = FALSE, warning = FALSE}
# Create a vector of variables 
DC2010_sf1_race_vars <- c(White = "P005003", 
              Black = "P005004", 
              Asian = "P005006", 
              Hispanic = "P004003")

# Retrieve variables and geographies for census block groups in Multnomah County, Oregon
bg_Mult_2010_dc <- tidycensus::get_decennial( # API call
  geography = 'tract', # Specify geography
  state = 'OR', # Specify state name
  county = c('Multnomah'), # Specify county name
  variables = DC2010_sf1_race_vars, # Specify the vector of variables
  year = 2010, # Specify year
  geometry = T, # Include or omit spatial geometry
  output = 'tidy', # Set to wide or tidy/long format
  summary_var = "P001001" # Specify summary variable (e.g., total population)
)

bg_Mult_2010_dc %>%
  mutate(pct = 100 * (value / summary_value)) %>% # Calculate the race/ethnicity proportion of the total population
  ggplot(aes(fill = pct)) + # Specify the aesthetic as fill based on the proportion values
  facet_wrap(~variable) + # Facet wrap on race/ethnicity
  geom_sf(color = NA) + # Specify outline color
  coord_sf(crs = 26910) + # Specify coordinate system/projection (e.g., UTM10N)
  ggtitle("Multnomah County", subtitle = "Population by Race/Ethnicity") + # Specify title and subtitle
  scale_fill_viridis_c() # Specify fill symbology
```

Create an interactive map using `leaflet`.

```{r class.source = 'fold-show', message = FALSE, warning = FALSE}
# Use leaflet's color ramp creator
medinc_colors <- leaflet::colorQuantile(
  palette = 'Greens', # This will accept any ColorBrewer color pallette
  domain = bg_Mult_2018_acs5$estimate, # Specify the variable
  n = 9 # Specify the number of breaks
)

co_2010_dc_WGS84 <- st_transform(co_2010_dc, 4326)
bg_Mult_2018_acs5_WGS84 <- st_transform(bg_Mult_2018_acs5, 4326)

leaflet() %>% 
  addTiles('http://{s}.tile.stamen.com/toner/{z}/{x}/{y}.png') %>% # Specify base map
  addPolygons( # Create map layer
    data = filter(co_2010_dc_WGS84, NAME == "Multnomah County, Oregon"), # Specify data set and query single observation
    weight = 3, # Specify outline width
    fillOpacity = 0, # Specify fill opacity
    color = 'royalblue', # Specify outline color
    group = "Counties (2010 DC)", # Specify legend item name
    popup = paste0( # Create pop-up
      "<b>2010 County</b> ",
      paste0("<br><b>Name</b>: ", co_2010_dc_WGS84$NAME, "<br><b>Total Population</b>: ", prettyNum(co_2010_dc_WGS84$pop_tot, ','))
    )
  ) %>%
  addPolygons(
    data = bg_Mult_2018_acs5_WGS84, # Create map layer
    weight = 1, # Specify outline width
    color = 'black', # Specify outline color
    fillColor = medinc_colors(bg_Mult_2018_acs5_WGS84$estimate), # Use created color ramp
    fillOpacity = 0.7, # Specify fill opacity
    group = "Block Groups (2018 ACS 5-year)", # Specify legend item name
    popup = paste0( # Create pop-up
      "<b>2018 ACS 5-year Block Group</b> ",
      paste0("<br><b>Median HH Income Estimate</b>: ", prettyNum(bg_Mult_2018_acs5_WGS84$estimate, ','), "<br><b>Median HH Income MOE</b>: ", prettyNum(bg_Mult_2018_acs5_WGS84$moe, ','))
    )
  ) %>%
  # Layers control
  addLayersControl(
    overlayGroups = c("Counties (2010 DC)", "Block Group (2018 ACS 5-year)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

# Export and Import

Write out data to CSV, Excel, RDS, Esri feature classes, shapefiles, etc.

```{r warning=FALSE, error=FALSE, message=FALSE, class.source = 'fold-show', results='hide'}
# write_csv(DC2010_sf1, "./DC2010_sf1.csv") # From `readr` package by Hadley Wickham; not the same as write.csv (former is twice as fast and does not write rows names/extra Field 1 column)
# write.xlsx(st_2010_dc, "./st_2010_dc.xlsx") # From `xlsx` package by Adrian Dragulescu
# saveRDS(bg_Mult_2018_acs5,"./bg_Mult_2018_acs5.rds")
# st_write(bg_Mult_2010_dc,"./bg_Mult_2010_dc.shp")
# arc.write(path = "./R.gdb/co_OR_2018_acs5", data = co_OR_2018_acs5, shape_info = list(type = 'Polygon', WKT = 'GEOGCS["GCS_North_American_1983",DATUM["D_North_American_1983",SPHEROID["GRS_1980",6378137,298.257222101]],PRIMEM["Greenwich",0],UNIT["Degree",0.017453292519943295]]'), overwrite = T)
```

Read in data from various formats to avoid duplicative calls to Census API.

```{r warning=FALSE, error=FALSE, message=FALSE, class.source = 'fold-show', results='hide'}
# csv <- read_csv("./DC2010_sf1.csv")
# xlsx1 <- read_excel("./st_2010_dc.xlsx", sheet = "sheet_name") # from the tidyverse
# xlsx2 <- read.xlsx("./st_2010_dc.xlsx", sheetName = "county_births_wide")
# bg_Mult_2018_acs5_RDS <- readRDS("./bg_Mult_2018_acs5.rds")
# bg_Mult_2010_dc  <- st_read("./bg_Mult_2010_dc.shp")
```
